Heavy rainfall in Paraguay during the 2015-2016 austral summer: causes and sub-seasonal-to-seasonal predictive skill

### 2018-03-21 reports/writeup/paper.tex
Edits from reviewer comments

At seasonal timescales, \gls{enso} is the dominant driver of convection variability in the \gls{lprb}.
During El Ni\~no years, a low-level anticyclonic anomaly over central Brazil enhances occurrence of the low-level jet, favoring the development of mesoscale convective systems \citep{Velasco1987}.
The intensity and precise extent of this anomaly is relevant for the precise impact of \gls{enso} events.
The region also exhibits substantial variability between seasons of rainfall during El Ni\~no years, including a reversal of rainfall anomalies between November of that year and January of the following one, influenced by land-surface interactions \citep{Grimm2003,Grimm2009}.
Even beyond El Ni\~no years, regional land-surface feedbacks can cause regions that exhibit wet anomalies in the spring to experience more summer precipitation on average \citep{Grimm2007}.
Similarly, mid-latitude dynamics influence low-level wind anomalies on many time scales, though this relationship is complicated due to coupled tropical-extratropical interactions \citep{Jones2002,Carvalho2004}.
To address these potential interactions, a cross-timescale approach based on synoptic circulation types is employed here to diagnose the causes of the rainfall events; this method has been used in previous work for southeastern South America \citep{Munoz2015,Munoz2016} and other regions \citep{Moron2015}.

We begin by describing our data sources in \cref{sec:data} and our methods in \cref{sec:methods}.
In \cref{sec:diagnostics} we start our diagnosis highlighting the observed flooding and contextualizing it within a long river stage time series; we then use composites and a weather typing analysis to diagnose the circulation patterns associated with the heavy rainfall during NDJF 2015-16.
We turn in \cref{sec:fcsts} to the question of whether the observed rainfall was successfully predicted by available models.
To carry out this analysis we study both forecasts targeting the entire series for a limited area, and also forecasts targeting a large spatial area for only the first week of December, when the most important flooding events started.
We also explore the impact on forecasts of several bias-correction schemes.
In \cref{sec:discussion} we discuss limitations and potential implications of our findings and potential future work, and we present our concluding remarks in \cref{sec:summary}.

@@

This study analyzes probabilistic seasonal and sub-seasonal forecasts of heavy rainfall events, defined by exceedance of the \nth{90} percentile of modeled rainfall.

The seasonal predictions used  are known as ``flexible format'' forecasts, provided by the \gls{iri}.
These forecasts use a multi-model ensemble approach, with bias-corrected retrospective probabilistic forecasts produced using a total of (a) 144 members forced by evolving sea-surface temperatures, and (b) 68 members forced by persisted sea-surface temperatures; for details see \citet{Barnston2010}.
Flexible format means that the user of these forecasts can arbitrarily choose  particular thresholds (percentiles) to compute the  probability of exceedance (or non-exceedance) from the complete  probability density function of the climatological distribution,  rather than using the more common tercile categories.
The DJF 2015-2016 forecasts analyzed were produced in  November 2015.
Due to the short sample of flexible format forecasts available (only for 2012-2016 at the time of writing this paper), no verification was performed for these seasonal predictions.
These forecasts are provided at a horizontal  resolution of \SI{2.5}{\degree}.
The DJF 2015-2016 forecasts analyzed were produced in November 2015.

@@

The $K$-means algorithm is guaranteed to converge to a local minimum of inter-cluster variance; to select the best partition, 500 simulations were created using the implementation in Python's \texttt{scikit-learn} package \citep{Pedregosa2012} and then the classifiability index of \citet{Michelangeli1995} was computed between each partition and the 499 others.
The partition whose classifiability index, averaged for all 499 pairwise comparisons, was the highest was selected.
Calculation of the classifiability index for several values of $K$ (supplemental figure S2) suggests that states with $K=5,6,\ldots,8$ are all reasonable.
We chose the solution $K=6$ because the clusters identified are qualitatively similar to those determined over southeastern South America \citep{Munoz2015,Munoz2016} and have an intuitive physical meaning, discussed further in the following sections.
We refer to the resulting clusters as \glspl{wt}.
From a physical point of view, the $K$-means algorithm helps identify typical atmospheric circulation patterns in the \gls{eof}-filtered field via clustering of days with similar streamfunction configurations.
These clusters can also be understood as proxies  of the available states of the system, or the most frequently visited trajectories in the phase space of the physical system \citep{Munoz2015,Munoz2016a,Munoz2017}.

@@

Logistic regression models the conditional probability of binary events, and it has been widely used in \gls{mos} approaches for binary outcomes.
Nonetheless, when using logistic regression to address multiple thresholds via independent fits, the predicted probabilities are, in general, not mutually consistent \citep{Messner2014}.
The \gls{xlr} was designed to address this shortcoming \citep{Wilks2009}, via the consideration of a transformation of the thresholds of interest as an additional predictor variable.
The \gls{hxlr}, a generalization of the \gls{xlr}, was proposed to appropriately use the ensemble spread as predictor for the dispersion of the predictive distribution \citep{Messner2014}.

@@

To understand how circulation anomalies observed during NDJF 2015-2016 led to the observed floods it is helpful to first explore the atmospheric circulations which are typically associated with heavy rainfall in the lower Paraguay River.

@@

There are advantages in simultaneously considering useful climate information at multiple timescales, rather than just focusing on one of them.
This multistage prediction system is known as the \gls{iri}'s ``Ready-Set-Go" approach \citep{Hellmuth2011,Goddard2014}.
For the sake of concision, in this section we discuss only probabilistic seasonal (DJF 2015-2016) and sub-seasonal (Dec 1-7, 2015) forecasts.

@@

\Cref{fig:chiclet} uses a Chiclet diagram \citep{Carbin2016} to visualize, as a function of lead time, the time evolution of the uncorrected, ensemble-mean rainfall anomaly forecast, spatially averaged over the \gls{lprb}.
At times greater than about two weeks, the ensemble-mean forecast is slightly wetter than climatology.
At weather timescales (less than one week), the ensemble-mean successfully predicts the timing and amplitude of the area-averaged rainfall.
At timescales of one to three weeks, the ensemble average successfully forecast the strongest breaks and pauses in the rainfall, such as the heavy rainfall during December 2015 and the dry period during mid-January 2016.

@@

This Atlantic-Pacific interaction may also be a source of dynamical model biases in the representation of rainfall events in the region.
In order to adequately forecast rainfall in certain parts of southeastern South America during El Ni\~no years, models need to reproduce not one but two stationary wave trains, originating the Pacific and the Atlantic, and their interactions \citep{Barreiro2017}.
The frontal activity is associated with the extension, shape and variability of the \gls{sacz}, and could be blocked by the presence of anomalous anti-cyclones off the coast of Brazil, like the ones described in reference to ``No-Chaco'' jet events.
Other mechanisms that have been known to modulate rainfall signals in this region include the \gls{sacz} \citep{Carvalho2004,Munoz2015,Munoz2016} and land-biosphere-atmosphere interactions \citep{Grimm2000,Grimm2007} which also tend to be poorly represented in models \citep{Koster2004,Green2017}.

>>>>

At seasonal timescales, \gls{enso} is the dominant driver of convection variability in the \gls{lprb}.
During El Ni\~no years, a low-level anticyclonic anomaly over central Brazil enhances occurrence of the low-level jet, favoring the development of mesoscale convective systems \citep{Velasco1987}.
The intensity and precise extent of this anomaly is relevant for the precise impact of \gls{enso} events.
The region also exhibits substantial variability between seasons of rainfall during El Ni\~no years, including a reversal of rainfall anomalies between November of that year and January of the following one, influenced by land-surface interactions \citep{Grimm2003,Grimm2009}.
Even beyond El Ni\~no years, regional land-surface feedbacks can cause regions that exhibit wet anomalies in the spring to experience more summer precipitation on average \citep{Grimm2007}.
Similarly, mid-latitude dynamics influence low-level wind anomalies on many time scales, though this relationship is complicated due to coupled tropical-extratropical interactions \citep{Jones2002,Carvalho2004}.
To address these potential interactions, a cross-timescale approach based on synoptic circulation types is employed here to diagnose the causes of the rainfall events.
This method has been used in previous work for southeastern South America \citep{Munoz2015,Munoz2016} and other regions \citep{Moron2015}.

The paper proceeds as follows.
We begin by describing our data sources in \cref{sec:data} and our methods in \cref{sec:methods}.
In \cref{sec:diagnostics} we start our diagnosis highlighting the observed flooding and contextualizing it within a long river stage time series; we then use composites and a weather typing analysis to diagnose the circulation patterns associated with the heavy rainfall during NDJF 2015-16.
We turn in \cref{sec:fcsts} to the question of whether the observed rainfall was successfully predicted by available models.
To carry out this analysis we study both forecasts targeting the entire series for a limited area, and also forecasts targeting a large spatial area for only the first week of December, when the most important flooding events started.
We also explore the impact on forecasts of several bias-correction schemes.
In \cref{sec:discussion} we discuss limitations and potential implications of our findings and potential future work, and we present our concluding remarks in \cref{sec:summary}.

@@

This study analyzes probabilistic seasonal and sub-seasonal forecasts of heavy rainfall events, which we define as exceedance of the \nth{90} percentile of NDJF daily precipitation across all ensemble members and initializations.

The seasonal predictions used  are known as ``flexible format'' forecasts, provided by the \gls{iri}.
These forecasts use a multi-model ensemble approach, with bias-corrected retrospective probabilistic forecasts produced using a total of 144 members forced by evolving sea-surface temperatures and 68 members forced by persisted sea-surface temperatures; for details see \citet{Barnston2010}.
Flexible format means that the user of these forecasts can arbitrarily choose  particular thresholds (percentiles) to compute the  probability of exceedance (or non-exceedance) from the complete  probability density function of the climatological distribution,  rather than using the more common tercile categories.
The DJF 2015-2016 forecasts analyzed were produced in  November 2015.
Due to the short sample of flexible format forecasts available (only for 2012-2016 at the time of writing this paper), no verification was performed for these seasonal predictions.
These forecasts are provided at a horizontal  resolution of \SI{2.5}{\degree}.
The DJF 2015-2016 forecasts analyzed were produced in November 2015.

@@

The $K$-means algorithm is guaranteed to converge to a local minimum of inter-cluster variance; to select the best partition, 500 simulations were created using the implementation in Python's \texttt{scikit-learn} package \citep{Pedregosa2012}.
Next, the classifiability index of \citet{Michelangeli1995} was computed between each partition and the 499 others.
The partition whose classifiability index, averaged for all 499 pairwise comparisons, was the highest was selected.
Calculation of the classifiability index for several values of $K$ (supplemental figure S2) suggests that states with $K=5,6,\ldots,8$ are all reasonable.
We chose the solution $K=6$ because the clusters identified are qualitatively similar to those determined over southeastern South America \citep{Munoz2015,Munoz2016} and have an intuitive physical meaning, which we discuss further in the following sections.
We refer to the resulting clusters as \glspl{wt}.
From a physical point of view, the $K$-means algorithm helps identify typical atmospheric circulation patterns in the \gls{eof}-filtered field via clustering of days with similar streamfunction configurations.
These clusters can also be understood as proxies  of the available states of the system, or the most frequently visited trajectories in the phase space of the physical system \citep{Munoz2015,Munoz2016a,Munoz2017}.

@@

Logistic regression models the conditional probability of binary events, given one or more predictors, and has been widely used in \gls{mos} approaches for binary outcomes.
Nonetheless, when using logistic regression to address multiple thresholds via independent fits, the predicted probabilities are, in general, not mutually consistent \citep{Messner2014}.
The \gls{xlr} was designed to address this shortcoming \citep{Wilks2009}, via the consideration of a transformation of the thresholds of interest as an additional predictor variable.
The \gls{hxlr}, a generalization of the \gls{xlr}, was proposed to appropriately use the ensemble spread as predictor for the dispersion of the predictive distribution \citep{Messner2014}.

@@

To understand how circulation anomalies observed during NDJF 2015-2016 led to the observed floods it is helpful to first explore the atmospheric circulations which are typically associated with heavy rainfall in the lower Paraguay River during the full observed record.

@@

There are advantages in simultaneously considering useful climate information at multiple timescales, rather than just focusing on one of them.
This multistage prediction system is known as the \gls{iri}'s ``Ready-Set-Go" approach \citep{Hellmuth2011,Goddard2014}.
In this study we examine only probabilistic seasonal (DJF 2015-2016) and sub-seasonal (Dec 1-7, 2015) forecasts.

@@

\Cref{fig:chiclet} uses a Chiclet diagram \citep{Carbin2016} to visualize, as a function of lead time, the time evolution of the uncorrected, ensemble-mean rainfall anomaly forecast, spatially averaged over the \gls{lprb}.
At times greater than about two weeks, the ensemble-mean forecast is for slightly positive rainfall anomalies at nearly all initialization dates and lead times.
At weather timescales (less than one week), the ensemble-mean successfully predicts the timing and amplitude of the area-averaged rainfall.
At timescales of one to three weeks, the ensemble average successfully forecast the strongest breaks and pauses in the rainfall, such as the heavy rainfall during December 2015 and the dry period during mid-January 2016.

@@

This Atlantic-Pacific interaction may also explain the skill of dynamical models for representing heavy rainfall in the region.
In order to adequately forecast rainfall in certain parts of southeastern South America during El Ni\~no years, models need to reproduce stationary wave trains originating in the Pacific and the Atlantic and their interactions \citep{Barreiro2017}.
Other mechanisms that have been known to modulate rainfall signals in this region include the \gls{sacz} \citep{Carvalho2004,Munoz2015,Munoz2016} and land-biosphere-atmosphere interactions \citep{Grimm2000,Grimm2007} which also tend to be poorly represented in models \citep{Koster2004,Green2017}.
The stationary wavetrain interactions, land-atmosphere interactions, and topography may explain why simulating heavy rainfall in this region is so difficult  \cref{fig:chiclet,fig:subs-prob-fcst}.
Improving understanding of these phenomena is important opportunity for \gls{s2s} prediction, and is left for future work.
